Next Trippy Structures\thanks{Supported by organization x.}

### 2019-06-10 paper.tex
Tirso Is Done

\begin{document}
%
\title{Next Trippy Structures\thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
With the increasing use of passenger tracking technology on public transportation networks, it is becoming necessary to store a constantly growing amount of information regarding the usage (demand) of the networks, which is different from the well-known information of stops and timetables (offer). To enable analysis over these large collections of data, a compact yet indexed representation is needed that, unlike the works that focus on indexing moving objects, allows for representation and querying of relevant network elements related to passenger trips.

@@

\subsection{Underlying structures and algorithms}
\subsubsection{Summed Area Tables}
The Summed Area Tables were first introduced in computer graphics \cite{crow1984summed} to speed up the mipmapping process.\marginpar{\tiny part of the explanation of \acumm~should go in here}

@@

Thus, we propose the T-Matrix, an accumulated solution that involves two-dimensional matrices of integers enabling aggregated queries by row, column, or window/range. In the context of a public trasnportation network, it would by useful to solve flexible line-centered queries; that is, queries that support aggregation by any dimension. Therefore, it could be possible to aggregate either by time-interval (e.g. number of users got on at any stop of the line on 2019/01/02); by stop (e.g. number of commuters that got on a bus at stop S); or by stop and time-interval.

In a naïve representation, it would only be necessary a two-dimensional matrix having stops as rows and journeys as columns, so each cell would have an integer representing the number of passengers that got on the vehicle at a specific stop in a particular time (journey). The main issue with this approach is that for each query it would be necessary to sum integer by integer each cell of the solution; hence, if we are trying to compute how many people got on a bus in several consecutive stops during a time interval (several consecutive journeys) it would be necessary to sum all the cells within the matrix that match these journeys and stops.

On the other hand, T-Matrix is based on this naive proposal but instead of storing each cell as a simple integer it assigns an accumulated value to each slot, being the top left cell the smaller number and the bottom right position the cumulative value resulting from adding all the previous cells. Despite having larger numbers in almost every slot of the matrix, this approach allows to apply the dynamic programing formula that follows:

@@

The combination of this operations with our proposed accumulative matrix enables to compute the same two-dimensional (or one-dimensional) queries in constant time O(1), as it is only neccesary to access and sum four cells in all the matrix.

EXAMPLE

One simple way of dealing with the growth of numbers in our cumulative solution is just apply a kind of sampling with differences; in this way, a basic example could be keep the middle column m ← (n s + 1)/2 explicitly, and representing the values in the other columns m±k as the difference with respect to column m.

Taking this simple structure one step further we have built a cumulative matrix with several sampling columns, reducing significantly the size of the structure. Hence, this new cumulativematrix is divided in blocks having n-factor sampling columns, calculating n-factor as a percentage
of the total number of columns. During the queries real numbers in the matrix are obtained by adding each cell plus the nearest sampling column on the left.

In order to reach a lighter solution we combined the two last compression techniques building a cumulative matrix sampled by columns which have a centered sampled column in turn. Thus, a value in a cell of the matrix would be obtained adding its value to the value of the nearest sampling column on the left which would be calculted by summing this sampled value to the center column.

Due to comparison reasons it has been also built these sampled structures using sampling rows instead of sampling columns as it was explained, leading this to the the seven structures that have been used in the experiments of this work: the basic matrix (BASIC), the cumulative matrix (SUM), the cumulative matrix with a central sampling column as reference (DIFF), the cumulative matrix with several sampling columns that divide it in blocks (COL-BLOCK), the COL-BLOCK matrix with a second central sampling (COL-BLOCK-DIFF) and the row version of the last two, ROW-BLOCK AND ROW-BLOCK-DIFF respectively.

@@

\begin{itemize}
    \item \texttt{board\_X$_{LT}$} This can be solved easily in the \acumm, as there are aggregated matrices for boarding stops, even though it is necessary to access a separate matrix for every line queried, hence $O(l)$. For \ttctr, the number of occurrences for the stop X is counted by delimiting its range on \FuncSty{D} (with a constant time $select$ operation) and filtering down the \texttt{WM} if needed. For \ctr, the filtering is done through \FuncSty{WML} and \FuncSty{WMJ}, and after that we must subtract the occurrences of final stops, obtained by one \texttt{end\_X$_{LT}$} with the same restrictions.
    \item \texttt{alight\_X$_{LT}$} Both in \ttctr~and \ctr, the only alighting stops that are explicitly represented are the final stops. To obtain an approximated count of the rest of them, we would need to extract and examine all the occurrences for \texttt{board\_W$_{LT}$} for every W that could have been boarded before X, as the worst-case time complexity of solving with their index operations is prohibitive. On the other hand, it can be solved much faster in \acumm~by accessing the alighting matrices.
    \item \texttt{board\_L$_T$} Straightforward in \acumm, as it only needs to access one matrix. For \ttctr~we must filter through the \texttt{WM} for every stop from L. While in \ctr~the total number of occurrences of line L can be calculated in one $O(\log|L|)$ operation (with additional filtering through \FuncSty{WMJ} if needed), we need to subtract the occurrences of start\_L$_T$ and end\_L$_T$, the later having a greater complexity of $O(|S| \times \ctrq)$.
    \item \texttt{alight\_L$_T$} Can be approximated both in \ttctr~and \ctr~with board\_L$_T$, given that every user that boards on a line is expected to alight from it within a reasonable time.
    \item \texttt{board\_T} and \texttt{alight\_T} Although \acumm, \ttctr~and \ctr~could solve it by applying board\_L$_T$ $|L|$ times, this would imply a prohibitive cost for both \ttctr~and \ctr, where extracting all the trips would be preferred in practice.
    \item \texttt{load\_X$_{LT}$} This can be easily calculated in \acumm~as detailed in Section~\ref{sec:acumm}. \marginpar{Tirso please explain how!}
    \item \texttt{start\_X$_{LT}$} Cannot be solved in \acumm. In the two other representations, we need to the range for the pattern $\$X$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed.
    \item \texttt{end\_X$_{LT}$} In \ctr~it is solved similarly to \texttt{start\_X$_{LT}$}, by delimiting the range for the pattern $X\$$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed. It can be much more complex for \ttctr, where the only way to restrict a line is to make a new query for every stop that could have been boarded before X in that line, which is a problem detailed at the start of Section~\ref{sec:ctr}.
    \item \texttt{switch\_X$_{LT}$} Calculated as board\_X$_{LT}$ - start\_X$_{LT}$.
    \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} Explained in detail in Algorithm~\ref{alg:xy}. In \ttctr~the $|S|$ factor only takes effect if the line (or time) for Y must be restricted, as for \texttt{end\_X$_{LT}$}.
    \item \texttt{start\_L$_T$} Trivially solved in \ctr, by filtering down \FuncSty{WML} and \FuncSty{WMJ}, if needed, over the $\$$ section of the \FuncSty{CSA}. On the other hand, for \ttctr~the \texttt{CSA} needs to be queried with the $\$X$ pattern for every stop X from line $l$.
    \item \texttt{end\_L$_T$} Can be solved in \ctr, by performing a end\_X$_{LT}$ for every stop X from the line $l$. If we did the same for \ttctr, we would end up with a complexity of $O(|S|^2\times\ttctrq)$, which we consider prohibitive.
    \item \texttt{start\_T} Solved in the \ctr~by applying \texttt{start\_X$_{LT}$} $|L|$ times.
    \item \texttt{end\_T} None of our representations can solve this query efficiently, although it could be approximated by \texttt{start\_T} in CTR.
\end{itemize}

@@

We have implemented the most adequate queries for \ttctr~and \ctr~from those described in Section~\ref{sec:rq}, and measured their average execution time from 100.000 randomly generated queries on [rellenar detalles de Compostela 2]. In this section we will only discuss four of the possible configurations tested for both \ttctr~and \ctr, with t$_{\Psi}$ values of 32 and 512 combining with \texttt{WM} using either uncompressed bitvectors (\texttt{RG32}) or the most compressed (\texttt{RRR128}). These four configurations should be illustrative enough to provide an understanding of the space-time trade-offs of our approaches, although the execution times for all tested configurations can be found in the Appendix A.

>>>>

\begin{document}
%
\title{Next Trippy Structures\thanks{Funded in part by European Union`s Horizon 2020 research and innovation programme
under the Marie Sklodowska-Curie grant agreement No 690941 (project
BIRDS). The Spanish group is also partially funded by Xunta de Galicia/FEDERUE
[CSI: ED431G/01; GRC: ED431C 2017/58; GEMA:IN852A
2018/14 and Innovapeme: IN848D-2017-2350417]; by MINECO-AEI/FEDER-UE
[Datos 4.0: TIN2016-78011-C4-1-R; and ETOMERDFD3:
TIN2015-69951-R]; and by FPI Program [BES-C-2017-0085]. M. A. Rodr´ıguez is
partially funded by Fondecyt-Conicyt grant number 1170497 and by the Millennium
Institute for Foundational Research on Data. }}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Nieves R. Brisaboa\inst{1} \and
Antonio Fari\~na\inst{1}\and
Daniil Galaktionov\inst{1,2} \and Tirso V. Rodeiro\inst{1} \and M. Andrea Rodr\'iguez\inst{3,4}}
%
\authorrunning{N. Brisaboa et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidade da Coru\~na, Laboratorio de Bases de Datos, A Coru\~na, Spain  \email{\{brisaboa, fari, d.galaktionov, tirso.varela.rodeiro\}@udc.es}  \and 
Enxenio S.L., A Coru\~na, Spain
\and
Computer Science Department, Universidad de Concepci\'on, Concepci\'on, Chile \and Millennium Institute for Foundational Research on Data, Santiago, Chile
\email{andrea@udec.cl }\\
 }
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

\keywords{Compact structures  \and Compression \and Trajectories}
\end{abstract}
%
%
%
\section{Introduction}
With the increasing use of passenger tracking technology on public transportation networks, it is becoming necessary to store a constantly growing amount of information regarding the usage (demand) of the networks, which is different from the well-known information of stops and timetables (offer). To enable analysis over these large collections of data, a compact yet indexed representation is needed that, unlike the works that focus on indexing moving objects, allows for representation and querying of relevant network elements related to passenger trips.

@@

\subsection{Underlying structures and algorithms}
\subsubsection{Summed Area Tables}
The Summed Area Tables were first introduced in computer graphics \cite{crow1984summed} to speed up the mipmapping process. In this work we use this approach in order to manage aggregated data about user trajectories based on the idea that an amount of commuters share the same ride (with all its implications: times, trajectories...).

@@

\begin{figure}[h!]
  \includegraphics[scale=0.65]{Tmatrices.png}
  \caption{T-Matrices example}
  \label{fig:tmatrix}
\end{figure}

Thus, we propose the T-Matrix, an accumulated solution that involves two-dimensional matrices of integers enabling aggregated queries by row, column, or window/range. In the context of a public transportation network, it would be useful to solve flexible line-centered queries; that is, queries that support aggregation by any dimension. Therefore, it could be possible to aggregate either by time-interval (e.g. number of users got on at any stop of the line on 2019/01/02); by stop (e.g. number of commuters that got on a bus at stop S); or by stop and time-interval.

In a na\"{\i}ve representation, it would only be necessary a two-dimensional matrix having stops as rows and journeys as columns, so each cell would have an integer representing the number of passengers that got on the vehicle at a specific stop in a particular time (journey). The main issue with this approach is that for each query it would be necessary to sum integer by integer each cell of the solution; hence, if we are trying to compute how many people got on a bus in several consecutive stops during a time interval (several consecutive journeys) it would be necessary to sum all the cells within the submatrix that match these journeys and stops.

On the other hand, T-Matrix is based on this na\"{\i}ve proposal but instead of storing each cell as a simple integer it assigns an accumulated value to each slot, being the top left cell the smaller number and the bottom right position the cumulative value resulting from adding all the previous cells. Despite having larger numbers in almost every slot of the matrix, this approach allows to apply the dynamic programming formula that follows:

@@

The combination of this operations with our proposed accumulative matrix enables to compute the same two-dimensional (or one-dimensional) queries in constant time O(1), as it is only necessary to access and sum four cells in all the matrix.

There are many options for compressing this cumulative matrix in order to get smaller dimensions, in this work we considered two relatively simple options with competitive results. One simple way of dealing with the growth of numbers in our cumulative solution is just apply a kind of sampling with differences; in this way, a basic example could be keep the middle column (\textit{$\mathsf{middle $\leftarrow$  (  \mathopen|c\mathclose|   + 1)/2} $}) explicitly, and representing the values in the other columns \textit{m±k} as the difference with respect to column \textit{m} (Diff in figure \ref{fig:tmatrix}). Being the algorithm to revert this differences as the one shown in Algorithm  \ref{alg:undiff}.

\begin{algorithm}[hbt!]

\SetKwData{value}{value}
 \SetKwFunction{extract}{Undiff}
 \SetKwProg{Fn}{Function}{\string:}{}

 \Fn{\extract{x, y}}{
 \KwData{coordinates x , y} 
 \KwResult{Value in the original cumulative matrix}
 \BlankLine
 \value $\leftarrow$ 0\;
 \BlankLine

\eIf{$ y == middle $}
            {$ \value = diffMatrix[x][y]$}
            {\eIf{$ y < middle $}
                {$ \value = diffMatrix[x][middle] - diffMatrix[x][y]$}
                {$ \value = diffMatrix[x][y] - diffMatrix[x][middle]$}}

 \BlankLine
 \Return{\value}\;
 }

 \caption{Obtaining the real value of any coordinates in the original cumulative matrix through the difference matrix}
 \label{alg:undiff}
\end{algorithm}

Taking this simple algorithm one step further we have built a cumulative matrix with several sampling rows, reducing significantly the size of the structure. Hence, this new differential matrix (Blocks in figure \ref{fig:tmatrix}) is divided in square blocks where the first line in each block remains the same as in the cumulative matrix while the rest of the rows in the block are calculated from it. A more technical description would be as follows in the simplified algorithm \ref{alg:unblock}.

\begin{algorithm}[hbt!]

\SetKwData{value}{value}
 \SetKwFunction{extract}{Unblock}
 \SetKwProg{Fn}{Function}{\string:}{}

 \Fn{\extract{x, y, block\_size}}{
 \KwData{coordinates x , y; 
         block_size: dimension of the each block} 
 \KwResult{Value in the original cumulative matrix}
 \BlankLine
 \value $\leftarrow$ 0\;
 \BlankLine

\eIf{$ x  \mod  block\_size $}
            {$ \value = blockMatrix[x][y]$}
            {$ \value = blockMatrix[x][y] + blockMatrix[x - (x  \mod  block\_size)][y]$}

 \BlankLine
 \Return{\value}\;
 }

 \caption{Obtaining the real value of any coordinates in the original cumulative matrix throught the difference matrix}
 \label{alg:unblock}
\end{algorithm}

@@

\begin{itemize}
    \item \texttt{board\_X$_{LT}$} This can be solved easily in the \acumm, as there are aggregated matrices for boarding stops, even though it is necessary to access a separate matrix for every line queried, hence $O(l)$. For \ttctr, the number of occurrences for the stop X is counted by delimiting its range on \FuncSty{D} (with a constant time $select$ operation) and filtering down the \texttt{WM} if needed. For \ctr, the filtering is done through \FuncSty{WML} and \FuncSty{WMJ}, and after that we must subtract the occurrences of final stops, obtained by one \texttt{end\_X$_{LT}$} with the same restrictions.
    \item \texttt{alight\_X$_{LT}$} Both in \ttctr~and \ctr, the only alighting stops that are explicitly represented are the final stops. To obtain an approximated count of the rest of them, we would need to extract and examine all the occurrences for \texttt{board\_W$_{LT}$} for every W that could have been boarded before X, as the worst-case time complexity of solving with their index operations is prohibitive. On the other hand, it can be solved much faster in \acumm~by accessing the alighting matrices.
    \item \texttt{board\_L$_T$} Straightforward in \acumm, as it only needs to access one matrix. For \ttctr~we must filter through the \texttt{WM} for every stop from L. While in \ctr~the total number of occurrences of line L can be calculated in one $O(\log|L|)$ operation (with additional filtering through \FuncSty{WMJ} if needed), we need to subtract the occurrences of start\_L$_T$ and end\_L$_T$, the later having a greater complexity of $O(|S| \times \ctrq)$.
    \item \texttt{alight\_L$_T$} Can be approximated both in \ttctr~and \ctr~with board\_L$_T$, given that every user that boards on a line is expected to alight from it within a reasonable time.
    \item \texttt{board\_T} and \texttt{alight\_T} Although \acumm, \ttctr~and \ctr~could solve it by applying board\_L$_T$ $|L|$ times, this would imply a prohibitive cost for both \ttctr~and \ctr, where extracting all the trips would be preferred in practice.
    \item \texttt{load\_X$_{LT}$} It is possible to solve this kind of query using both, alighting and boarding matrices. Knowing how many commuters got on and off the vehicle previously to one particular point makes trivial to determine how many of them are in the vehicle at each tract. This method is easily adapted to measure the average number through an interval of time.
    \item \texttt{start\_X$_{LT}$} Cannot be solved in \acumm. In the two other representations, we need to the range for the pattern $\$X$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed.
    \item \texttt{end\_X$_{LT}$} In \ctr~it is solved similarly to \texttt{start\_X$_{LT}$}, by delimiting the range for the pattern $X\$$ in the \FuncSty{CSA} and filtering down \FuncSty{WML} and \FuncSty{WMJ} if needed. It can be much more complex for \ttctr, where the only way to restrict a line is to make a new query for every stop that could have been boarded before X in that line, which is a problem detailed at the start of Section~\ref{sec:ctr}.
    \item \texttt{switch\_X$_{LT}$} Calculated as board\_X$_{LT}$ - start\_X$_{LT}$.
    \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} Explained in detail in Algorithm~\ref{alg:xy}. In \ttctr~the $|S|$ factor only takes effect if the line (or time) for Y must be restricted, as for \texttt{end\_X$_{LT}$}.
    \item \texttt{start\_L$_T$} Trivially solved in \ctr, by filtering down \FuncSty{WML} and \FuncSty{WMJ}, if needed, over the $\$$ section of the \FuncSty{CSA}. On the other hand, for \ttctr~the \texttt{CSA} needs to be queried with the $\$X$ pattern for every stop X from line $l$.
    \item \texttt{end\_L$_T$} Can be solved in \ctr, by performing a end\_X$_{LT}$ for every stop X from the line $l$. If we did the same for \ttctr, we would end up with a complexity of $O(|S|^2\times\ttctrq)$, which we consider prohibitive.
    \item \texttt{start\_T} Solved in the \ctr~by applying \texttt{start\_X$_{LT}$} $|L|$ times.
    \item \texttt{end\_T} None of our representations can solve this query efficiently, although it could be approximated by \texttt{start\_T} in CTR.
\end{itemize}

@@

We have implemented the most adequate queries for \ttctr~and \ctr~from those described in Section~\ref{sec:rq}, and measured their average execution time from 100.000 randomly generated queries on a Intel Xeon E5-2620v4@2.1 GHz machine. In this section we will only discuss four of the possible configurations tested for both \ttctr~and \ctr, with t$_{\Psi}$ values of 32 and 512 combining with \texttt{WM} using either uncompressed bitvectors (\texttt{RG32}) or the most compressed (\texttt{RRR128}). These four configurations should be illustrative enough to provide an understanding of the space-time trade-offs of our approaches, although the execution times for all tested configurations can be found in the Appendix A.
